Red Wine Quality Exploration by Layali R.
========================================================


```{r echo=FALSE, packages}
library(ggplot2)
library(corrplot)
```

```{r echo=FALSE, Load_the_Data}
# Load the Data
df = read.csv("C:/users/layal/Desktop/R Project/wineQualityReds.csv")
```

> This dataset is about Red Wine components. The goal is to identify the main components of wine, the relationships between these components and develop an idea about what distinguishes good from bad wine.

# General analysis
> Please note that below I combined the bivariate and multivariate plot sections because they are tightly connected, I found that a story will be told better if I immediately followed a bivariate plot with a related multivariate plot.

```{r echo=FALSE, General_analysis}
dim(df)
str(df)
summary(df)
```
> Check if there are missing values

```{r echo=FALSE}
print("Is there any missing values?")
ifelse(nrow(subset(df,is.na(df))) != 0, "Yes", "No")
```

> There are no missing values.

> Check if observations are unique, I will remove the column marking the
sequence number for each observations, it is not needed in my EDA

```{r echo=FALSE, message=FALSE, warning=FALSE}
df$X <- NULL
print("Are all observations are unique?")
ifelse(nrow(unique(df)) == nrow(df), "Yes", "No")
```

> Yes! All observations are unique.

> I will add two columns to dataset, total acidity and quality bucket. Later in this report I will describe how I reach the conclusion that total acidity should include citric acid and fixed acidity. Also, quality bucket is going to help in the multivariate analysis.

```{r echo=FALSE, message=FALSE, warning=FALSE}
df$total.acidity = df$fixed.acidity+df$citric.acid
df$quality.bucket <- ifelse(df$quality<=5,"Low Quality", "High Quality")
```

# Univariate Plots Section

```{r echo=FALSE, message=FALSE, warning=FALSE}
ggplot(aes(x=quality), data = df) +
  geom_histogram()
  
table(df$quality)
```

> The quality of the red wine in this data set ranges between 3 and 8. There is very small number of wine with very low quality of 3 or very high quality of 8. The observations seem to be right skewed with short tail, they center around 5.5.

```{r echo=FALSE, message=FALSE, warning=FALSE}
ggplot(aes(x=fixed.acidity), data = df) +
  geom_density(alpha=0.2, fill="blue")

summary(df$fixed.acidity)
```

> Most of the observations have fixed.acidity of about 7. The distribution is right skewed with a tail expanding until about 16. Too much fixed acidity will result in tartar wine, this explains why there are small number of observations with high amounts of this acid.
There are no values of zero or 1 so it should be safe to try the log scale.

```{r echo=FALSE, message=FALSE, warning=FALSE}
ggplot(aes(x=log10(fixed.acidity)), data = df) +
  geom_density(alpha=0.2, fill="red") +
  geom_vline(aes(xintercept=mean(log10(fixed.acidity), na.rm=T)),
             color="red", linetype="dashed", size=1)
summary(log10(df$fixed.acidity))
```

> The density plot of the log10 of this variable looks "close" to normal the mean is 0.91 and is represented using the vertical red line, it is not quite at the center because the density is not normal.

```{r echo=FALSE, message=FALSE, warning=FALSE}
ggplot(aes(x=volatile.acidity), data = df) +
  geom_density(alpha=0.2, fill = "yellow")
```

> The density plot is bimodal. Volatile acidity is found in very small amounts in red wine, it appears when wine is steamed. More about this variable later in this report.

```{r echo=FALSE, message=FALSE, warning=FALSE}
ggplot(aes(citric.acid), data = df) + 
  geom_histogram(binwidth = 0.005)
sort(table(df$citric.acid))
```

> Citric acids usualy appear in citris fruit, they give freshness to the wine. The above histogram of citrict.acid does not show any particular pattern.
The majoriy of the observations does not include any amount of this acid.

```{r echo=FALSE, message=FALSE, warning=FALSE}

ggplot(aes(residual.sugar), data = df) + 
  geom_density(fill="green")
summary(df$residual.sugar)
```

> The density looks right skewed with long tail. Even when transformed to log10 (not interesting enough to include in this document), it also looks right skewed. Log10 is not useful because most values are very small. The description document mentions that "wines with greater than 45 grams/liter are considered sweet". There are no sweet wines in this sample. The maximum value is 15.5.

```{r echo=FALSE, message=FALSE, warning=FALSE}
ggplot(aes(chlorides), data = df) + 
  geom_density(fill="orange")
```

> Similar observations can be seen when plotting the chlorides. Most observations have very small amounts of this compound. It is clearly right skewed. Chlorides are present in wine from materials used to sterilize wine making equipments, detergents are not commonly used to clean the  equipments because they leave residue. Based on this reasoning, I don't think chloride is major component of wine.

```{r echo=FALSE, message=FALSE, warning=FALSE}
ggplot(aes(free.sulfur.dioxide), data = df) + 
  geom_density(alpha = 0.2, fill = 'blue')

ggplot(aes(total.sulfur.dioxide - free.sulfur.dioxide), data = df) + 
  geom_density(alpha = 0.2, fill = 'blue') +
  xlab("bound sulfur")

summary(df$free.sulfur.dioxide)

ggplot(aes(total.sulfur.dioxide), data = df) + 
  geom_density(alpha = 0.2, fill = 'blue')
```

> According to the dataset description, the total.sulfur.dioxide is a superset of free.sulfur.dioxide and bound form of SO2. Most observations have small concentration of free or bound SO2. Only few observations have relatively higher concentrations of these materials which result in right skewed density plot.

```{r echo=FALSE, message=FALSE, warning=FALSE}
ggplot(aes(sulphates), data = df) +
  geom_density()
```

> The majority of observations have relatively small amount of Sulphate. Even at small amounts suphur can be antioxidant, anti aging and hence, it preserves the freshness of the wine. The density plot is right skewed. Sulphate contribute to SO2 substances so it will be interesting to study these materials together.

```{r echo=FALSE, message=FALSE, warning=FALSE}
ggplot(aes(density), data = df) + 
  geom_density() +
  geom_vline(aes(xintercept=mean(density, na.rm=T)),
           color="red", linetype="dashed", size=1)
mean(df$density)
sd(df$density)
```

> Density is normally distibuted with center around the mean of  0.9967 and sd of 0.0019.
Density depends on alcohol and sugar percentage so it should interesting to study these three variables together later in this report.

```{r echo=FALSE, message=FALSE, warning=FALSE}
ggplot(aes(pH), data = df) +
  geom_density() +
  geom_vline(aes(xintercept=mean(pH, na.rm=T)),
             color="red", linetype="dashed", size=1)
mean(df$pH)
sd(df$pH)
```

> pH measures the concentration of fixed acidity and citric acidity. Volatile acidity is measured using steam distillation process. Higher amounts of acid lead to lower values of pH and more acidic taste. pH is log scaled by nature. pH is normally distributed  with a center of 3.3, and sd of 0.15. pH of red wine ranges between 3 and 4, which is the case for this dataset. There are 29 exceptions with a pH value lower than 3. The majority of these exceptions are of quality 3 and 4.

```{r echo=FALSE, message=FALSE, warning=FALSE}
d <- subset(df, (df$pH < 3))
table(d$quality)
```

```{r echo=FALSE, message=FALSE, warning=FALSE}
ggplot(aes(alcohol), data = df) +
  geom_density()
```

> I don't think this is bimodal and it is probably a stretch to just call it right skewed. Most wines have about 9.5% alcohol, few observations have higher percentages of alcohol with a maximum of 15%.

```{r echo=FALSE, message=FALSE, warning=FALSE}
tail(names(sort(table(df$alcohol))), 1)
summary(df$alcohol)
```

# Univariate Analysis

> What is the structure of your dataset?
The dataset has 1599 observations and 13 numerical variables.

> What is/are the main feature(s) of interest in your dataset?
Quality is the main feature of interest.

> What other features in the dataset do you think will help support your \
investigation into your feature(s) of interest?
The main question to study is which factors affecting wine quality. In addition, I deduce some relationships between components when I read the dataset description. These relationships are (1) acidity and pH, (2) density vs alcohol and residual sugar, (3) sulphates and sulfur dioxides, (4) total sulphates and free sulphates. I look forward to investigating these relationships, either by verifying or debunking them.
Furthermore, I read about red wine in http://winemakersacademy.com and would
like to apply what I learnt there in this EDA.

> Did you create any new variables from existing variables in the dataset?
Quality bucket to distiguish between high and low quality wines.
Also, total.acidity which is the sum of citric and fixed acidity.

> Of the features you investigated, were there any unusual distributions? \
Did you perform any operations on the data to tidy, adjust, or change the form \
of the data? If so, why did you do this?
Most of the relationships were right skewed with long or short tail. There are two normally distributed relationships, density and pH. Volatile.acidity has a bimodal distribution. Furthermore, I couldn't find a pattern for citric acidity.


# Bivariate Plots Section

> I have 13 variables so I will start by plotting the correlation coefficient

```{r echo=FALSE, message=FALSE, warning=FALSE}
new.df <- df
new.df$quality.bucket <- NULL
m = cor(new.df)
corrplot(m, method = 'circle')
```

> From the plot above, I think it will be useful to look at the following relationships:

> fixed.acidity vs citric.acid, fixed.acidity vs pH (negative), volatile.acidity vs citric.acid (negative), citric.acid vs pH (negative)

> quality vs alcohol, quality vs sulphates, quality vs citric.acid, quality vs. fixed.acidity, quality vs volatile.acidity (negative)

> density vs fixed acidity, density vs alcohol (negative), density vs citric.acid, density vs residual sugar, density vs pH (negative)

### Acidity (fixed, volatile and citric).

> Recall pH is the measurement of strength of acidity, and acidity is a measurement of how much acid you have. Usually, the higher the amound of acid the smaller the pH. Therefore, it is expected to see linear relationship between pH and acidity. Volatile acids are not measured by pH though. Although volatile acids seem to have a positive relationship with pH in the correlation plot above, it is likely a fake relationship that is created by other factors.

```{r echo=FALSE, message=FALSE, warning=FALSE}
ggplot(aes(x=pH, y = citric.acid), data = df) +
  geom_point() +
  geom_smooth(method="lm")

cor.test(df$pH,df$citric.acid)
```

> There is a good negative relationship between pH and fixed acidity. The pearson correlation coefficient is -0.54.
It it also interesting to see zero amount of citric acid and different values of pH, implying that pH depends on other substances as we will see below.

```{r echo=FALSE, message=FALSE, warning=FALSE}
ggplot(aes(x=pH, y = volatile.acidity), data = df) +
  geom_point(color='red') +
  geom_smooth(method="lm")
cor.test(df$pH,df$volatile.acidity)
```
> There is a weak relationship of 0.23 between volatile acidity and pH. The Pearson correlation coefficient is 0.24.
Again, I think this relationship is fake as pH is not an indication of volatile acidity.

```{r echo=FALSE, message=FALSE, warning=FALSE}
ggplot(aes(x=pH, y = fixed.acidity), data = df) +
  geom_point(color='brown') +
  geom_smooth(method="lm")
cor.test(df$pH,df$fixed.acidity)
```

> In high concentrations, the fixed acids can give the wine tartaric taste. There is strong negative relationship between pH and fixed acidity. The pearson correlation coeficient is -0.68. Fixed acidity is very important in wine industry, it gives each wine its own characteristic. Fixed and citric acids are the two factors that contribute to pH in this dataset.

```{r echo=FALSE, message=FALSE, warning=FALSE}
ggplot(aes(x=pH, y = fixed.acidity, color=citric.acid), data = df) +
  geom_point() +
  geom_smooth(method="lm")
cor.test(df$pH,(df$fixed.acidity+df$citric.acid))
```

> The plot above shows that when you have high fixed and citric acids amounts the pH will be low and vice versa. The sum of these two substances have a - 0.69 correlation coefficient. Based on this analysis, I added a column of total acidity to the dataset, this variables is the sum of fixed and citric acids.

### Density

> The dataset description says that the main components of density is sugar and alcohol, I will test the accuracy of this statement.

```{r echo=FALSE, message=FALSE, warning=FALSE}
ggplot(aes(x=density, y = alcohol), data = df) + 
  geom_point(color = 'pink') +
  ylim(8,13) +
  geom_smooth(method='lm')
cor.test(df$density, df$alcohol)
```

> There is a negative relationship with an R of -0.50 between density and alcohol.

```{r echo=FALSE, message=FALSE, warning=FALSE}
ggplot(aes(x=density, y = residual.sugar), data = df) + 
  geom_jitter(alpha = 1/5) +
  ylim(0.5, 5) +
  geom_smooth(method='lm')
cor.test(df$density, df$residual.sugar)
```

> There is a positive weak relationship with as R of 0.36 between density and residual.sugar.

```{r echo=FALSE, message=FALSE, warning=FALSE}
ggplot(aes(x=density, y = total.acidity), data = df) + 
  geom_jitter(alpha = 1/5) +
  geom_smooth(method='lm')
cor.test(df$density, df$total.acidity)
```

> There is a good positive relationship with an R of 0.66 between denstiy and total.acidity.

> I conclude that residual sugar doesn't correlate much with density, rather alcohol and total acidity seems to be the major contributers. I think more statistical analysis are needed to determine if the relationship between density and total acidity is real or an artifact of other relationships, but I cannot see that by using my current tools

```{r echo=FALSE, message=FALSE, warning=FALSE}
ggplot(aes(x=density, y = total.acidity, color = alcohol), data = df) + 
  geom_point() +
  geom_smooth(method='lm', color='pink')
```

> This plot shows, again, the nice positive relationship between density, and total acidity and a negative one with alcohol.


### Quality

> Note that the lack of observations with quality of 3, 4, 7 and 8 makes it hard to make conclusions out of the plots below.

```{r echo=FALSE, message=FALSE, warning=FALSE}
ggplot(aes(x=as.factor(quality), y = alcohol), data = df) + 
  geom_boxplot()
```

> Quality increases with alcohol, the boxplot shows this with the exception of quality 5 which has lower mean of alcohol with respect to quality of 4. Quality 5 also has a large number of outliers. There is a positive relationship between alcohol and the wine quality with R of 0.48.

```{r echo=FALSE, message=FALSE, warning=FALSE}
ggplot(aes(x=as.factor(quality), y = volatile.acidity), data = df) + 
  geom_boxplot()
```

> As for the volatile acidity, the mean of alcohol of each quality category is getting smaller with increasing quality. The negative relationship can also be shown with R value of -0.39.

> Quality depends weakly on sulphates (R = 0.25), citric.acid (R = 0.23) and total.acidity (R = 0.17), as shown below.

```{r echo=FALSE, message=FALSE, warning=FALSE}
ggplot(aes(y=sulphates, x = as.factor(quality)), data = df) + 
  geom_boxplot()
```

> This is a weak positive relationship with the quartiles (almost) increasing with quality and the mean of sulphates for each category also increases. The R value for this relationship is 0.25.

```{r echo=FALSE, message=FALSE, warning=FALSE}
ggplot(aes(y=citric.acid, x = as.factor(quality)), data = df) + 
  geom_boxplot()
```

> Although we can see the positive relationship between citric acid and quality, the small number of observations in category 3 is making it hard to make conclusions because the huge quartile of category 3 is probably weakening the relationship between citric acid and quality. The R value is 0.23.

```{r echo=FALSE, message=FALSE, warning=FALSE}
ggplot(aes(y=fixed.acidity, x = as.factor(quality)), data = df) + 
  geom_boxplot()
```

> Fixed acidity also has a weak relationship with quality, qualities of 3 and 8 and the too few observations is affecting the increasing trend. This relationship has an R value of 0.17.

```{r echo=FALSE, message=FALSE, warning=FALSE}

ggplot(aes(x=alcohol, y = volatile.acidity), data = df) +
  geom_jitter(color = 'red', alpha = 1/5) +
  ylim(0.125, 1.2) +
  facet_wrap(~df$quality.bucket)
ggplot(aes(x=volatile.acidity, fill = quality.bucket), data = df) +
  geom_density(alpha=0.2)

```

> The plots above show other views of the major two factors contributing to quality. The first plot shows the low quality win (<=5) vs high quality (> 5). I solved overplotting by adding a jitter. The weak relationships do not make it easy to view the relationship but one can notice that high quality has relatively smaller values of volatile acidity and higher numbers of alcohol percentages. Second plot is interesting because it "kind of" breaks down the bimodal density plot in volatile acidity, it turns out that the higher mod belongs to low quality wine.

```{r echo=FALSE, message=FALSE, warning=FALSE}
ggplot(aes(x=alcohol, y = sulphates,
           color = quality.bucket), data = df) +
  geom_jitter(alpha=1/5) +
  ylim(0.3,1.3)
```

> Recall, that low quality in this plot refers to quality of less than or equal to 5, otherwise the wine is in the high category. It is interesting to see that low quality wine has smaller amount of alcohol and also smaller amounts of sulphates.

> Other potential relationships I will investigate next are: sulphates and chlorides, free, bound, total sulfur dioxide and sulphates.

### Sulphates and chlorides

```{r echo=FALSE, message=FALSE, warning=FALSE}
ggplot(aes(x=sulphates, y = chlorides), data = df) +
  geom_jitter(alpha=1/10) +
  ylim(0, 0.2) +
  xlim(0.2, 1.25) +
  geom_smooth(method='lm')
cor.test(df$sulphates, df$chlorides)
```

> There is a weak relationship between sulphates and chlorides with R = 0.37. I'm not a chemist but I looked it up and there is no chemical relationship between the two compounds so I think that this relationship is fake.


### Free, bound, total sulfur dioxide and sulphates

```{r echo=FALSE, message=FALSE, warning=FALSE}
ggplot(aes(x=free.sulfur.dioxide, y= total.sulfur.dioxide),
       data = df) +
  ylim(0, 150) +
  xlim(0, 60) +
  geom_jitter(alpha = 1/10, color = "purple")
```

> Recall that total sulfur dioxide is a superset of free.sulfur.dioxide and bound sulfur dioxide (per the dataset description). The plot confirms this because we don't see a point where free sulfur dioxide is larger than the total.
It seems like bound sulfur dioxide composed most of the total sulfur dioxide because the range of sulphates is small.

```{r echo=FALSE, message=FALSE, warning=FALSE}
ggplot(aes(x=sulphates, y= total.sulfur.dioxide),
       data = df) +
    ylim(0, 150) +
    xlim(0.25, 1.5) +
  geom_jitter(alpha = 1/10, color = "red") +
  geom_smooth(method="lm")

```

> The plot also shows that the majority of observations have small sulphates value mainly between 0.4 to 0.8. It also shows a number of observations with high total.sulfur.dioxide value and still very low sulphate value. Therefore, there is no obvious linear relationship between the two variables although their description implies a relationship. I plot the smooth line and it confirms my expectations.

# Bivariate Analysis

> Talk about some of the relationships you observed in this part of the \
investigation. How did the feature(s) of interest vary with other features in \the dataset?
I found a positive relatively good relationship between wine quality and alcohol, and a negative weak relationship with volatile acidity. Wine quality depends on other factors but the main components seem to be alcohol and volatile acidity.

> Did you observe any interesting relationships between the other features (not the main feature(s) of interest)?
(1) There is a relatively strong negative relationship between fixed acidity and pH. Also a weaker relationship exists between citric acidity and pH. As for volatile acidity there is a weak negative relationship that I suspect to be fake due to some related documents I read about how the acidity density is measured in win.
(2) Analysis reveals good negative relationship between density and alcohol. However, the relationship with residual sugar seems to be weaker and positive. Rather, I found a stronger positive relationship between density and total acidity (fixed acidity + citric acidity).
(3) Dataset description claims a relationship between sulphates and sulfur dioxides but I couldn't find any worthy relationship between the two variables.
(4) Total sulphates consists of free sulphates and bound sulphates. I found that total sulphates are superset of free sulphates and it looks like the majority of sulphates is of type bound.

> What was the strongest relationship you found?
The relationship between pH and total acidity has the strongest negative R value of 0.69.


# Multivariate Analysis

> Talk about some of the relationships you observed in this part of the \
Contrary to what the dataset description says, the relationship between residual sugar and density is weaker than density relationship with total acidity (R of 0.66). However, alcohol which is also mentioned in the description has a good R of -0.50.

> Were there any interesting or surprising interactions between features?
The fixed acidity and citric acidity have an interesting strong relationship with pH.

------

# Final Plots and Summary

```{r echo=FALSE}
maplabel <- function(var) {
  result <- paste("Quality", as.character(var), sep=" ")
  return(result)
}
```

### Plot One

```{r echo=FALSE, message=FALSE, warning=FALSE, Plot_One}
ggplot(aes(x=alcohol, y = volatile.acidity), data = df) +
  geom_jitter(color = 'red', alpha = 1/5) +
  ylim(0.125, 1.2) +
  ylab("Volatile Acidity (g / dm^3)") +
  xlab("Alcohol Percentage")
  facet_wrap(~df$quality.bucket, scales="free")
```

### Description One

> Low Quality refers to quality equal or less than 5 while High Quality is when quality is greater than 5. The plot shows the main two factors affecting the quality of red wine according to this dataset; volatile acidity and alcohol. The more alcohol percentage the better wine, however, volatile acidity seem to lower the quality of red wine when it is increased.

### Plot Two

```{r echo=FALSE, message=FALSE, warning=FALSE, Plot_Two}
ggplot(aes(x=density, y = total.acidity, color = alcohol), data = df) + 
  geom_point() +
  xlab("Density (g/cm^3)") +
  ylab("Total Acidity (g / dm^3)")
  geom_smooth(method='lm', color='pink')
```

### Description Two

>  This plot shows the nice positive relationship between density and total acidity and a negative one with alcohol. Residual sugar doesn't correlate much with density as the dataset description says. It is rather alcohol and total acidity that seem to be the major contributers. I think more statistical analysis are needed to determine if the relationship between density and total acidity is real or an artifact of other relationships, but I cannot see that by using my current tools.

### Plot Three

```{r echo=FALSE, message=FALSE, warning=FALSE, Plot_Three}
ggplot(aes(x=pH, y = fixed.acidity, color=citric.acid), data = df) +
  geom_point() +
  geom_smooth(method="lm")
```

### Description Three

> In high concentration the fixed acids can give the wine tartaric taste. There is strong negative relationship between pH and fixed acidity. The pearson correlation coeficient is -0.68. Fixed acidity is very important in wine industry, it gives each wine its own characteristic. Fixed and citric acids are the two factors that contribute to pH in this dataset.

## Reflection

> This EDA was interesting, I proved a few relationships, debunked some and found new ones. I enjoyed learning about red wine components. I wish there were more observations, especially of low and high qualities so some relationships would become clearer. It is well polished dataset with all unique observations and no missing values.
I also felt that I need to develop stronger chemistry background to better understand the relationships between the different compounds.
I wasn't suprised to find that alcohol is a major contributer to good wine, but volatile acidity was a little surprising to stand out in a negative relationship, I expected total acidity to be a stronger contributer.